**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-08-07 submission/manuscript.Rmd
update after meeting with Pat 3

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs is large, then that would indicate that the models were overfit. difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.

To evaluate the risk for over-optimism of each model, we calculated the range for each model using 100 splits. The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models. If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`. These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. Therefore, it is important to employ the hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.

__Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017]. Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies. ML models decrease in interpretability as they increase in complexity. In this study we used two methods to interpret models with varying complexity.

We interpreted linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the absolute feature weights of the trained models.  We ranked the absolute values of the weights of all the OTUs for each data-split [Figure 3]. We calculated the median ranks of these features over the 100 data-splits. In the three linear models, the OTUs that had the largest median rank and drove the detection of SRNs contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU00822, OTU00768), Clostridium (OTU00609), Blautia (OTU00659, OTU00012), Faecalibacterium (OTU00015), Gamella (OTU00426) and Peptostreptococcus (OTU00367) [Figure 3]. Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that … were shared among both of the L2-regularized models.

We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a method called permutation importance on the held-out test data.  Permutation importance analysis is a posthoc explanation of the model where we randomly permute non-correlated features individually and groups of highly correlated features together. We then calculate how much the predictive performance of the model (i.e AUROC values) decrease when each OTU or group of OTUs is permuted randomly. We ranked the OTUs based on how much they decreased the median testing AUROC; the OTU with the largest decrease ranking highest. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure 4]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].

To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3]. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings based on weights [Figure 3] and permutation importance [Figure S3]. Similarly, L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively. However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities. Collinearity in a microbial dataset is when one OTU is dependant on another OTU. The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret models using feature weights. Our interpretation of L2-regularized logistic regression based on feature weight rankings, showed that an OTU that belongs to *Lachnospiraceae* (OTU00056) was not among the most important OTUs in making a prediction [Figure 3C]. However, it was picked out as important when the same model was interpreted using permutation importance [Figure S3C]. This is due to collinearity in the dataset, where *Lachnospiraceae* (OTU00056) is significantly correlated with *Feacalibacterium* (OTU00015) which has the highest ranked feature weights [Figure 3C]. We need to investigate the relationships among features to identify the true underlying factors when making a prediction.

@@

We compared the training times of the seven ML models. As the complexity of a ML model and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5]. Linear models trained faster than non-linear models. L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression had the shortest training times with `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, a decision tree, SVM with radial basis function kernel, random forest and XGBoost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

@@

>>>>

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.  Subsample experiment results to show what overfitting looks like goes here…

To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 splits. The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` on average across the seven models. If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`. These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. Therefore, it is important to employ the hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.

__Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017]. Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies. ML models decrease in interpretability as they increase in complexity. In this study we used two methods to help interpret our models.

We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median absolute feature weights for each OTU in the models [Figure 3]. The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) and Peptostreptococcus (OTU 367) [Figure 3]. Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that 4 OTUs were shared among the five highest ranked OTUs for both of the L2-regularized models.

We interpreted the feature importance in the non-linear models using permutation importance. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest. The OTUs that had the largest impact and drove the detection of SRNs across the non-linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629), Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58), Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477), Parvimonas (OTU 361), Roseburia (OTU 822) [Figure 4]. Among the twenty OTUs with the largest impact across these four models, there was only one OTU (OTU 822) that was shared among all of the models; however, among the tree-based models, we found three OTUs (OTU 367, OTU 58 and OTU 110) to be consistently important. Similarly, in the random forest and XGBoost models, four of the twenty most important OTUs were shared between the models.

When we analyzed the L1-regularized SVM with linear kernel model using permutation importance, the approach picked out many of the same OTUs (e.g. OTU00822, OTU01239, OTU00609) that were important when we interpreted the model using feature rankings based on weights [Figure 3]. Similarly, L2-regularized SVM and L2-regularized logistic regression  each identified many OTUs that were selected by both interpretation methods.

Interestingly, for the all the models, except the poorly performing L1-regularized SVM and decision tree, the variation in the impact of these important OTUs was minimal. This supports the findings of others that a classification requires numerous OTUs, rather the presence of a single or a few of the OTUs [@sze_leveraging_2018].

@@

We compared the training times of the seven ML models. As expected, the training times increased as the complexity of a ML model and the number of tuned hyperparameter settings increased and the linear models trained faster than non-linear models [Figures S1-S2; Figure 5]. When we subsetted the size of the training dataset, we observed a [linear, quadratic, etc] relationship between the size of the dataset and the training time for the random forest model.

@@

Overview
•	Establish pipeline that can be generalized to any modeling method
•	Linear/non-linear not a bfd
•	Interpretability tradeoff
Point 1
•	Importance for transparency, picking the most complex model is not necessarilty the best model
•	Good / better / best:  Table S1
•	“best” has multiple meanings
Point 2
•	Our paper is concerned with developing models to gain greater biological insights into what’s going on
•	Still need biological validation and further experimentation
Point 3
•	These models show great opportunity to use microbiome to make diagnoses
•	We talked about classification, but others have used it in microbiome data to do regression, mulit-class – pipeline is the same
•	Others have used this on shotgun metagenomic data and collinearity is expected to be an even larger problem since 5000 genes are all correlated with each other
•	Deployment is a different question that requires independent datasets to test on; we currently lack decent sized parallel datasets where the same type of sequence data and clinical data have been collected
•	We also have a problem with OTU-based approaches
Conclusion/Future direction
•	ML is awesome and there’s always new methods; our framework gives structure to investigators wanting to try these out. We encourage reseraches to apply multiple models to their data and report the results
•	Things like neural nets are attractive and hold great promise, but we need a ton more data to apply and they are impossible to interpret
